---
title: "2024_AlpineStrategy_distributionmodeling"
author: "Michelle DePrenger-Levin"
date: "2024-04-17"
output: html_document
---

```{r}

library(raster) ## says getData will be removed, use geodata instead
library(geodata)
## library(rgdal) # retired in 2023. move to sf/stars/terra with GDAL and PROJ
# library(maptools) # retired in 2023; same for rgeos; no longer active maintainer

library(sf)
library(devtools)
devtools::install_github("rsbivand/sp@evolution")
Sys.setenv("_SP_EVOLUTION_STATUS_"=2)  ## so that sp will call sf::st_crs() to validate coordinate reference system

library(prism)
library(ggplot2)
library(distributional)

```


World Climate data
 standard (19) WorldClim Bioclimatic variables for WorldClim version 2. They are the average for the years 1970-2000.

    BIO1 = Annual Mean Temperature  
    BIO2 = Mean Diurnal Range (Mean of monthly (max temp - min temp))  
    BIO3 = Isothermality (BIO2/BIO7) (* 100)  
    BIO4 = Temperature Seasonality (standard deviation *100)  
    BIO5 = Max Temperature of Warmest Month  
    BIO6 = Min Temperature of Coldest Month  
    BIO7 = Temperature Annual Range (BIO5-BIO6)   
    BIO8 = Mean Temperature of Wettest Quarter  
    BIO9 = Mean Temperature of Driest Quarter  
    BIO10 = Mean Temperature of Warmest Quarter  
    BIO11 = Mean Temperature of Coldest Quarter  
    BIO12 = Annual Precipitation  
    BIO13 = Precipitation of Wettest Month  
    BIO14 = Precipitation of Driest Month  
    BIO15 = Precipitation Seasonality (Coefficient of Variation)  
    BIO16 = Precipitation of Wettest Quarter  
    BIO17 = Precipitation of Driest Quarter  
    BIO18 = Precipitation of Warmest Quarter  
    BIO19 = Precipitation of Coldest Quarter   
    
    
This is WorldClim version 2.1 climate data for 1970-2000. This version was released in January 2020.       
Variable	Description	Unit    

    tmin	minimum temperature	°C    
    tmax	maximum temperature	°C    
    tavg	average temperature	°C    
    prec	total precipitation	mm    
    srad	incident solar radiation	kJ m-2 day-1    
    wind	wind speed (2 m above the ground)	m s-1   
    vapr	vapor pressure	kPa   


```{r}

# bioclim_data <- raster::getData("worldclim", var = 'bio', res = 0.5, lon = 105, lat = 40)
bioclim_data <- geodata::worldclim_tile('bio', res = 0.5, lon = 105, lat = 40, path=tempdir())

bioclim_data

plot(bioclim_data)

gadm_data <- raster::getData("GADM", country = "USA", level = 2)


```

