---
title: "2024_AlpineStrategy_distributionmodeling"
author: "Michelle DePrenger-Levin"
date: "2024-04-17"
output: html_document
---


Predict species distributions from iNat and herbarium collections. Make detection histories and apply site-occupancy to estimate 'true' species distributions corrected for imperfect detection.    

```{r}

library(raster) ## says getData will be removed, use geodata instead
library(geodata)
library(dismo)
## library(rgdal) # retired in 2023. move to sf/stars/terra with GDAL and PROJ
# library(maptools) # retired in 2023; same for rgeos; no longer active maintainer

library(terra)
library(maps)
library(sf)
library(devtools)
devtools::install_github("rsbivand/sp@evolution")
Sys.setenv("_SP_EVOLUTION_STATUS_"=2)  ## so that sp will call sf::st_crs() to validate coordinate reference system

library(prism)
library(ggplot2)
library(usmap)
library(tidyterra)

library(mapview)
library(mapedit)
library(tidyverse)


library(dplyr)
library(tidyr)


## JAGS models
library(R2jags)
library(runjags)
library(mcmcplots)
library(boot)
```


World Climate data
 standard (19) WorldClim Bioclimatic variables for WorldClim version 2. They are the average for the years 1970-2000.

    BIO1 = Annual Mean Temperature  
    BIO2 = Mean Diurnal Range (Mean of monthly (max temp - min temp))  
    BIO3 = Isothermality (BIO2/BIO7) (* 100)  
    BIO4 = Temperature Seasonality (standard deviation *100)  
    BIO5 = Max Temperature of Warmest Month  
    BIO6 = Min Temperature of Coldest Month  
    BIO7 = Temperature Annual Range (BIO5-BIO6)   
    BIO8 = Mean Temperature of Wettest Quarter  
    BIO9 = Mean Temperature of Driest Quarter  
    BIO10 = Mean Temperature of Warmest Quarter  
    BIO11 = Mean Temperature of Coldest Quarter  
    BIO12 = Annual Precipitation  
    BIO13 = Precipitation of Wettest Month  
    BIO14 = Precipitation of Driest Month  
    BIO15 = Precipitation Seasonality (Coefficient of Variation)  
    BIO16 = Precipitation of Wettest Quarter  
    BIO17 = Precipitation of Driest Quarter  
    BIO18 = Precipitation of Warmest Quarter  
    BIO19 = Precipitation of Coldest Quarter   
    
    
This is WorldClim version 2.1 climate data for 1970-2000. This version was released in January 2020.       
Variable	Description	Unit    

    tmin	minimum temperature	°C    
    tmax	maximum temperature	°C    
    tavg	average temperature	°C    
    prec	total precipitation	mm    
    srad	incident solar radiation	kJ m-2 day-1    
    wind	wind speed (2 m above the ground)	m s-1   
    vapr	vapor pressure	kPa   


Not same extent, not sure if correct once changed with terra::ext
```{r}

# bioclim_data <- raster::getData("worldclim", var = 'bio', res = 0.5, lon = 105, lat = 40)
bioclim_data <- geodata::worldclim_tile('bio', res = 0.5, lon = -105, lat = 40, path=tempdir())
gadm_data <- geodata::gadm(country = "USA",  path=tempdir(), level = 2)

state <- map_data("state")
ggplot(data = state, aes(x = long, y = lat, fill = region, group = group))+
  geom_polygon(color = "white")+ 
  guides(fill=FALSE) + 
  theme(axis.title.x=element_blank(), axis.text.x=element_blank(), axis.ticks.x=element_blank(),
        axis.title.y=element_blank(), axis.text.y=element_blank(), axis.ticks.y=element_blank()) + 
  ggtitle('U.S. Map with States') + 
  coord_fixed(1.3)

plot(bioclim_data[[1]])
plot(gadm_data, add = TRUE)

CO_NM <- subset(state, region %in% c("colorado","new mexico"))
counties <- map_data("county")
CONM_counties <- subset(counties, region %in% c("colorado","new mexico"))

ggplot() +
  geom_spatraster(data = bioclim_data[[1]], aes(fill = tile_15_wc2.1_30s_bio_1))+
  geom_spatvector(data = gadm_data) 
# +
#   coord_cartesian(xlim = c(-110, -103), ylim = c(30,50))

```


Download GBIF records   
    
    Includes 'research grade' iNat observations
```{r}
phal <- sp_occurrence("Physaria", "alpina*", download = TRUE)

### Data cleaning
## Should exclude uncertainty of greater than 5000 meters
hist(phal$coordinateUncertaintyInMeters)
phal.clean <- phal %>%
  filter(coordinateUncertaintyInMeters < 30)


## create features object from gbif data. 
phal.sf <- sf::st_as_sf(phal.clean,
                        coords = c("lon","lat"),
                        crs = sf::st_crs(4326))

## Make hexagonal grid with res = 0.5 degrees in WGS 84
grid <- sf::st_make_grid(conm,
                         cellsize = 0.05,
                         crs = 4326,
                         square = FALSE) |> sf::st_as_sf()

grid$cellnum <- 1:length(grid$x)

## selct objects from grid cells containing points from phal.sf 
grid_subset <- sf::st_filter(grid, phal.sf)
st_bbox(grid_subset)
grid_crop <- sf::st_crop(grid, st_bbox(grid_subset))

## extract values from prec by polygons from grid_subset using mean as aggregate
results <- terra::extract(bioclim_data, terra::vect(grid_subset), fun = mean)


## create a SpatialPointsDataFrame from the data.frame
sp::coordinates(phal.clean) <- ~ lon + lat
crs(phal.clean) <- "+proj=longlat"
conm <- gadm_data[gadm_data$NAME_1 %in% c("Colorado","New Mexico")]
ext(conm)
ext(phal.clean) # -107, 10 38 63
ext(bioclim_data) ## 90 120 30 60


bioclim_crop <- crop(bioclim_data, grid_subset)
plot(bioclim_crop[[1]])
plot(grid_crop, color = NA, add = TRUE)
plot(phal.clean, add = TRUE)

phal$eventDate

##Extract values from bioclim layers by polygons from grid_subset using mean as aggregate
annmeantemp <- terra::extract(bioclim_data[[1]], terra::vect(grid_subset), fun = mean)
bioclimdata <- terra::extract(bioclim_data, terra::vect(grid_subset), fun = mean)

## Extract cell numbers for each specimen record
gridnum <- sf::st_intersection(grid_subset, st_as_sf(phal.clean, coords = c("lat","lon")))
## Build this for a list of alpine species to determine true '0's when someone made an observation in the area but didn't make a target species observation compared to no effort where no observations made in that cell in that year. 
mrc <- table(gridnum$cellnum, gridnum$year)

## Need to add zeros when other species were found but not PhAl
mrc[mrc > 0] <- 1


bioclim_data_ext <- bioclim_data
ext(bioclim_data_ext) <- ext(conm) 


bioclim_crop <- crop(bioclim_data, conm, ext = TRUE)
#### 
plot(bioclim_crop[[4]], main = "Physaria alpina iNat and herbarium")
plot(gadm_data, add = TRUE)
# plot(gadm_data[gadm_data$NAME_1 %in% c("Colorado","New Mexico")],
#      main = "Physaria alpina iNat and herbarium")
points(phal.clean, pch = 16, col = "red")


# overlay a SpatVector
ggplot()+
  geom_spatraster(data = bioclim_crop[[1]]) +
  # facet_wrap(~lyr, ncol = 2)  +
  scale_fill_whitebox_c(
    palette = "muted",
    labels = scales::label_number(suffix = "º"),
    n.breaks = 12,
    guide = guide_legend(reverse = TRUE)
  ) +
  geom_spatvector(data = gadm_data[gadm_data$NAME_1 %in% c("Colorado")], aes(fill = NAME_1, alpha = 0.5))


terra::crs(gadm_data, describe = TRUE) ## EPSG 4326
terra::crs(bioclim_data, describe = TRUE) ## EPSG 4326, -180, 180, 90, -90


ggplot() +
  geom_spatvector(data = gadm_data[gadm_data$NAME_1 %in% c("Colorado")], aes(alpha = 0.05)) +
  geom_spatraster(data = bioclim_crop[[1]]) +
  # facet_wrap(~lyr, ncol = 2)  +
  scale_fill_whitebox_c(
    palette = "muted",
    labels = scales::label_number(suffix = "º"),
    n.breaks = 12,
    guide = guide_legend(reverse = TRUE)
  ) +
  geom_point(data = phal[phal$coordinateUncertaintyInMeters < 100,], aes(x= lon, y=lat)) 

```


# model the true occupancy state, and the detection probablity based on temperature and precipitation   
AIC for full model and compare to other models   
```{r}

## JAGS binomial regression 
modelSDM_occupancy <- 
  paste("
        model {
          # Likelihood:
            for(i in 1:nSites){
              # Biological model (occupancy)
              z[i] ~ dbern(psi)  # z = 1 if occupied
              
              # Observation model
              y[i] ~ dbin(p*z[i], n)   # dbin( probability, number of trials)
            }
            
            
        
            }}")


## Organize the data
jagsData <- list(y = )

```


